**Evaluation of machine learning methods for 16S rRNA gene data**

### 2019-06-05 submission/manuscript.Rmd
add some comments from meeting
intergrate jennas changes

Machine learning (ML) modeling of the human microbiome is used to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but reliable ML models are scarce in part due to the flawed modeling methods. Furthermore, using black box ML models has hindered the validation of microbial biomarkers. To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict the presence/absence of colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases). To show the effect of model selection, we assessed the accuracy, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost). The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (101.3 h) and hard to interpret. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained the fastest (12 min). This study showed that we should choose ML models based on our expectations of performance and interpretability as well as our  computational resources. It also established standards for modeling pipelines of microbiome-associated ML models.

@@

@@

Machine learning (ML) methods are effective at recognizing and highlighting patterns in complex microbial datasets. Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].  However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented [@thaiss_persistent_2016; dadkhah_gut_2019]. More notably, flawed ML practices are prevalent such as using ML pipelines where there is no seperate held-out test dataset to evaluate model performance, reporting few or only the best outcomes of different randomizations of cross-validation and showing a disregard for large differences between cross-validation and testing performances as well as large confidence intervals of testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017]. Moreover, there is a lack of discussion on why a particular ML model is utilized. Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018].

The lack of transparency on modeling methodology and model selection negatively impact model reproducibility and reliability. We need to strive toward better machine learning practices by (1) implementing consistent and reliable machine learning pipelines and (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.

@@

@@

@@

 -

@@

#### Consideration of possible weaknesses
 - What happens with imbalanced data

 - What happens with smaller datasets

 - Does the code work on any microbiome data?

@@

For L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels we tuned the __cost__ hyperparameter which determines the regularization strength where smaller values specify stronger regularization. For SVM with radial basis function kernel we also tuned __sigma__ hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary. For the decision tree model, we tuned the __depth of the tree__ where deeper the tree, the more splits it has. For random forest, we tuned the __number of features__ to consider when looking for the best tree split. For xgboost, we tuned for __learning rate__ and the __fraction of samples__ to be used for fitting the individual base learners.
Models were trained using the machine learning wrapper caret package (v.6.0.81) in R (v.3.5.0).

>>>>

Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce in part due to the flawed (suboptimal) modeling methods that call into question the validity of developed ML models. Furthermore, using black box ML models has hindered the validation of microbial biomarkers. To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict the presence/absence of colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases). To show the effect of model selection, we assessed the accuracy, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost). The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (101.3 h) and hard to interpret. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained the fastest (12 min). This study showed that we should choose ML models based on our expectations of performance and interpretability as well as our  computational resources. It also established standards for modeling pipelines of microbiome-associated ML models.

@@

- Without rigorous pipelines, we can't believe the machine learning models.

- To speed up progress, let's slow down and make it right.

- Let's define and start using good ML practices.

@@

Machine learning (ML) methods are effective at recognizing and highlighting patterns in complex microbial datasets. Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].  However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented [@thaiss_persistent_2016; dadkhah_gut_2019]. More notably, ML practices such as using ML pipelines where there is no seperate held-out test dataset to evaluate model performance and reporting few or only the best outcomes of different randomizations of cross-validation are commonly used. When there are seperate testing sets to evaluate model performance, there are large differences between cross-validation and testing performances as well as large confidence intervals of testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017]. Moreover, there is a lack of discussion on why a particular ML model is utilized. Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018]. The lack of transparency on modeling methodology and model selection negatively impact model reproducibility and reliability. We need to strive toward better machine learning practices by (1) implementing consistent and rigourous machine learning pipelines and (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.

@@

- decision tree has no rgularization.

@@

- Testing with each model will take the same amount of time but if you are training a model and optimize hyper-parameters, then you need to consider the computational burden. The computational efficiency of different models have been investigated before. Explain for each model. In tree based models, depth of tree, number of features, .. matter. L1 linear is very fast.

- But we can reduce time paralellizing different components of the pipeline. Here I have parallelized each datasplit but we can also parallelize cross-validation hyerp-parameter search for each hyper-parameter setting.

@@

 - There is variability if we don't do a full grid search for the hyper-parameters. We can give each model 15 minutes to train and see how well they do. So an L1 linear model will give you terible results with one and not the other C value as seen in figure S1. We have done a granular search first and then limited the possible settings to a range that made the more sense.

@@

#### Consideration of possible weaknesses
 - What happens with imbalanced data
 The metric you use as the discriminative performance becomes important. Don't use accuracy.
 AUROC is good and random at 0.5 always.

 - What happens with smaller datasets
 Generalization performance depends on size. 
 The more complex the model, more data you will need.
 Use linear models if you have little data.

 - Does the code work on any microbiome data?
 So if yu have a new sample, do you have the same OTUs? NO

@@

For L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels we tuned the __cost__ hyperparameter which determines the regularization strength where smaller values specify stronger regularization. For SVM with radial basis function kernel we also tuned __sigma__ hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary. For the decision tree model, we tuned the __depth of the tree__ where deeper the tree, the more splits it has. For random forest, we tuned the __number of features__ to consider when looking for the best tree split. For xgboost, we tuned for __learning rate__ and the __fraction of samples__ to be used for fitting the individual base learners.

Hyperparameter selection is very important. We start with a granular grid search by then narrow and refine. Range of the grid depends on the problem. We can use hyper-band to help us with our hyperparameter selection. If not a full grid search is done, there could be a lot of variability in testing performance.
Models were trained using the machine learning wrapper caret package (v.6.0.81) in R (v.3.5.0).
